---
title: 'Chapter 3 Data Input | Chapter 4 Data Frames '
author: "Qianqian Shan"
date: "May 20, 2017"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "therbookdata")

```

##Data Input from the Keyboard


```{r}
x <- scan()
```

Data input from files have variables called `fields` and rows called `cases`.

If the file name is forgotten, *file.choose()* could be used.
```{r eval=FALSE}
data <- read.table(file.choose(), header = TRUE)
```

* Data Input using `read.table`

```{r}
data <- read.table("yields.txt", header = TRUE)
head(data)
```
* `read.delim` can omit `header = T`
```{r}
data <- read.delim("yields.txt")
```

* Data input using a defined function `rt`
```{r}
rt <- function(x) read.table(paste(x, ".txt", sep = ''),header = TRUE)
data <- rt("yields")
head(data,2)
```

As the default behavior of `read.table` is to convert character to factors, we need to use `as.is` to specify the columns that should =not be converted to factors. 

```{r}
murder <- read.table("murders.txt", header = TRUE)
str(murder)
murder <- read.table("murders.txt", header = TRUE, as.is = "region")
str(murder)
```

Data input directly from the web 

```{r}
data2 <- read.table("http://www.bio.ic.ac.uk/research/mjcraw/therbook/data/cancer.txt", header = TRUE)
# URL stands for universal resource locator 
head(data2)
```


## Read data using `scan()`

```{r}
# scan will create a list of vectors and we'd like a data frame 
data <- as.data.frame(scan("worms.txt", skip = 1, what= as.list(rep("", 7)))) # skip=1 will skip the header, 
# the last argument specify seven fields of character variables  ""

# the variable names are long and meanlingness, next obtain the names and apply them to the data 
header <- unlist(scan("worms.txt", nlines = 1, what = as.list(rep("",7))))
header
names(data) <- header
head(data,2)
```


Input from complex file structures using `scan`

```{r}
sapply(1:5, function(i) as.numeric(na.omit(scan("rt.txt", sep = "\t", quiet = TRUE)[(4*i - 3): (4*i)]))) 
# quiet = T prevents the printing "Read 20 itmes"...
```

## Reading data from a file using `readLines`
This is an alternative of `scan`.

```{r}
line <- readLines("worms.txt")
line 
```
Strip out the tab "\\t" 


```{r}
db <- strsplit(line, "\t") # returns a set of lists 
db <- (unlist(db))
dim(db) <- c(7, 21) # variable names dimention comes first 
t(db)[-1, ] # the first row is the names 

# change it to data frame 
frame <- as.data.frame(t(db)[-1,])
head(frame)

# add names 
names(frame) <- t(db)[1, ]
head(frame)
```

Read non-standard files using `readLines`
```{r}
readLines("rt.txt")
```
Split first on tabs and then on lines 

```{r}
rows <- lapply(strsplit(readLines("rt.txt"), split = "\t"), as.numeric)
rows
strsplit(readLines("rt.txt"), split = "\n") # this is ONE string

# remove NAs from each of the vectors 
sapply(1:5, function(i) as.numeric(na.omit(rows[[i]])))
```

## Warnings when you `attach` the dataframe 
The best approach is NOT to use `attach` 

```{r}
murder <- read.table("murders.txt", header = TRUE, as.is = "region")
attach(murder) # warning message shows up as the dataframe name is identical with one of the variable names 
               # better to rename the dataframe name 
detach(murder)
data <- read.table("murders.txt", header = TRUE, as.is = "region")
attach(data)
detach(data)
```


Check files exists from the command line 
```{r}
file.exists("Decay.txt")
```

Read dates and times from file, **refer Chapter 2 for instance**

**file paths** 
* set working directory by `setwd`
* `basename` returns the last path of a complete path 
* `dirname` returns the full path except for the last 

```{r}
basename("c:/temp/thesis/data")
dirname("c:/temp/thesis/data")
file.path("","p1","p2","p3", c("file1", "file2"))
basename(file.path("","p1","p2","p3", c("file1", "file2")))
dirname(file.path("","p1","p2","p3","filename"))
```

# Chapter 4 : Dataframes 

A **dataframe** is an object with rows and columns. 
Ways to create a dataframe: 

* Use `read.table` to read fils .

* Use `data.frame` function to bind together a numner of vectors. 



```{r}
worms <- read.table("worms.txt", header = TRUE)
```



Summary of a dataframe 

* `summary` 
* `by` to summarize the dataframe on the basis of factor levels 
* `aggregate` 

```{r}
# by(worms[,c(2,3)], worms$Vegetation, sum)
by(worms, worms$Vegetation, function(x) lm(Worm.density ~ Soil.pH, data= x))
aggregate(worms[, c(2, 3, 5, 7)], by = list(veg = worms$Vegetation), mean)

# or with more than one classifying factors 
aggregate(worms[, c(2, 3, 5, 7)], by = list(veg = worms$Vegetation, d = worms$Damp), mean)
```


Note the *different* classes of these two: 

```{r}
class(worms[3, ])
class(worms[, 3])
```

Select rows from a dataframe randomly 

```{r}
worms[sample(1:20, 8, replace = FALSE), ]
# sample(x, size, replace = FALSE, prob = NULL)
```


## Sorting dataframes 

```{r}
worms[order(worms$Slope), ][1:3,]

# order reversely 
worms[rev(order(worms$Slope)), ][1:5, ]

# order by the first and ties broken by the second, third ... 

worms[order(worms$Vegetation, worms$Worm.density), ][1:5, ]

# select columns by variable names 
worms[order(worms$Vegetation, worms$Worm.density), c("Vegetation", "Worm.density", "Soil.pH", "Slope")][1:6, ]
```

## Using logical conditions to select rows from dataframe 

```{r}
worms[worms$Damp == TRUE, ]

# Use logical operator 
worms[worms$Worm.density > median(worms$Worm.density) & worms$Soil.pH < 5.2, ]

# extract all numeric columns 
sapply(worms, is.numeric)
worms[, sapply(worms, is.numeric)]

# similarly, extract all factor columns 
worms[, sapply(worms, is.factor)]

# exclude certain rows 
worms[!(worms$Vegetation == "Grassland"), ]

# or use which function 
worms[ - which(worms$Damp == FALSE), ]

# or 
worms[!(worms$Damp == FALSE), ]

# or 
worms[worms$Damp == TRUE, ]
```

## Omitting rows containint missing values `NA` 
* `na.omit` 
* `na.exclude` , similar with `na.omit`, but different in the class of `na.action`  attribute of the result, and thus `na.exclude` padded the original length by inserting `NA` for using `naresid` and `napredict`. 
* `complete.cases` returns logical vector indicating which cases are complete 

```{r}
data <- read.table("worms.missing.txt", header = TRUE)
dim(data)

nona <- na.omit(data)
dim(nona) # 3 NA values deleted 

nona1 <- na.exclude(data)
dim(nona1)

# 
complete.cases(data)
# Analogue of na.omit 
data[complete.cases(data), ]

# check the number of NA values of each column 
apply(data, 2, is.na)
# count the NA values 
apply((apply(data, 2, is.na)), 2, sum)
```

## Using `order` and `!duplicated` to eliminate pseudoreplication 

Extract each vegetation type and each has the highest density within that vegetation type.
```{r}
# order data by density 
new <- worms[rev(order(worms$Worm.density)), ]
new[!duplicated(new$Vegetation),]
``` 
## Complex ordering with mixed directions 
There may be multiple sorting variables with different sorting directions. 
```{r}
# sort Vegetation in alphabetical order and density in decreasing order 
worms[order(worms$Vegetation, -worms$Worm.density), ]

# As using minus sign only works for numerical variables, so for factors, we need to 
# first use "rank" to convert levels to numeric 
worms[order(-rank(worms$Vegetation), -worms$Worm.density), ]


# select columns that contains character "S"
grep("S", names(worms)) # returns the corresponding column number 
worms[, grep("S", names(worms))]
```

## A dataframe with row names instead of row numbers 
Can suppress the creation of row numbers and allocate unique names to each row by altering the syntax of the `read.table` function. For example, add `row.names= ` command. 
```{r}
worms2 <- read.table("worms.txt", header = TRUE, row.names = 1)
head(worms2) # row numbers are suppressed 
```

## Eliminating duplicated rows from a dataframe 
```{r}
dups <- read.table("dups.txt", header = TRUE)
dups # row 3 and 5 are the same 

# strip out duplicated rows 
unique(dups)  # row numbers are still the original ones 

# the row that are duplicates 
dups[duplicated(dups), ] 
```

## Dates in dataframes 

```{r}
nums <- read.table("sortdata.txt", header = TRUE)
head(nums, 3) # data is in format dmy  


# In order to order the data by date, first need to convert date into date time format 
# to avoid sorting based on day - month 
dates <- strptime(nums$date, format = "%d/%m/%Y")
nums <- cbind(nums, dates)
head(nums[order(dates), ])
``` 

## Using `match` function in dataframes 

```{r}
herb <- read.table("herbicides.txt", header = TRUE)
# add corresponding recommended 
recs <- data.frame(
  worms, hb = herb$Herbicide[match(worms$Vegetation, herb$Type)]
) 
# match returns a vector of the positions of (first) matches of its first argument in its second
head(recs)
``` 

## Merging two dataframes 

```{r}
lifeforms <- read.table("lifeforms.txt", header = TRUE)
flowering <- read.table("fltimes.txt", header = TRUE)
lifeforms
flowering 
# two data have species in common 

# merge with only rows had complete cases in both 
merge(flowering, lifeforms) 

# include all 
both <- merge(flowering, lifeforms, all = TRUE) # NA values produced 
both

# now add a new column from another data frame to the above data frame 
seeds <- read.table("seedwts.txt", header = TRUE)
head(seeds) # columns have different names 

merge(both, seeds, by.x = c("Genus", "species"), by.y = c("name1", "name2"))
```

## Adding margins to a dataframe 

```{r}
frame <- read.table("sales.txt", header = TRUE)
frame

# add row means 
people <- rowMeans(frame[, -1])
people <- people - mean(people)
new.frame <- cbind(frame, people)
new.frame

# add col mean 
season <- colMeans(frame[, -1])
season <- season - mean(season)  # cannot use rbind directly as columns are different 

# copy one row 
new.row <- new.frame[1, ]
new.row[1] <- "seasonal effects"
new.row[2:5] <- season
new.row[6] <- 0
new.frame <- rbind(new.frame, new.row)
new.frame

# use sweep to subtract the grand mean from each value 
gm <- mean(unlist(new.frame[1:5, 2:5])) # overall mean 
gm <- rep(gm, 4)
new.frame[1:5, 2:5] <- sweep(new.frame[1:5, 2:5], 2, gm) # sweep out summary statistic


# put the grand/ overall mean in the bottom right corner 
new.frame[6, 6] <- gm[1]
new.frame
```

