---
title: "Chapter 19 Mixed-Effects Models | Chapter 20 Non-Linear Regression"
author: "Qianqian Shan"
date: "June 9, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "therbookdata")
```

The essence in deciding whether a categorical variable should be treated as a fixed effect or random effect: 

1. Fixed effects influence only the **mean** of y.

2. Random effects influence only the **variance** of y. 


Random effects arise from two contrasting kinds of circumstances: 

1. observational studies with hierarchical structure, 

2. designed experiments with different spatial or temporal scales. 


**Fundamental assumptions of linear mixed-effects models**: 

1. Within group errors are independent with mean 0 , variance $\sigma^2$ and are independent of the random effects. 

2.  The random effects are normally distributed with mean 0 and covariance matrix $\Psi$. 

3. The random effects are independent in different groups. 

4. The covariance matrix doesn't depend on the group. 




## Replication and pseudo-replication 

**Replicates** properties: 

1. Independent 

2. NOT grouped together in one place as aggregation means that they are not spatially independent

3. Be of appropriate spatial scale 

4. Repeated measures are not replicates.


**Pseudo-replication** occurs when the data has more degrees of freedom than really has: 

1. temporal involving repeated measurements from the same individual 

2. spatial involving several measurements taken from the same vicinity. 



**Ways to deal with pseudo-replication**: 

1. Average away the pseudo-replication and carry out analysis on the means 

2. Carry out separate analysis for each time period 

3. Use proper time series analysis or mixed effects models. 


## `lme` and `lmer` functions 

* `lme` has separate fixed and random effects specification, $fixed = y \tilde{} 1$, $random = \tilde{} 1|a/b/c$, meaning that "there are three random effects with c nested within b and b is nested within a", $lme(fixed = y \tilde{} 1, randome = \tilde{}1|a/b/c)$. 

* `lmer` has fixed and random effects specified togher, $lmer(y \tilde{} 1 +(1|a/b/c)$.


## Best linear unbiased predictors 

In mixed-effects models, the correlation between the pseudo replicates within a group causes **shrinkage**, and the best linear unbiased predictor $a_i = (\overline{y_i} - \mu) (\frac{\sigma_a^2}{\sigma_a^2 + \sigma^2/n})$, where $\sigma^2$ is the residual variance, $\sigma_a^2$ is the between group variance which introduces the correlation between the pseudo replicates within each group. 

More details on BLUP later, or refer <https://dnett.github.io/S510/21BLUP.pdf>. 



## Designed experiments with different spatial scales: Split plots 

If we want to use `anova` to compare mixed models with different fixed effects structurs, we jmust use maximum likelihood: `method = "ML"` for `lme, `REML = FALSE` for `lmer`. 

For more details on `REML`, see <https://dnett.github.io/S510/20REML.pdf>. 

1. If the experiment is balanced and there are no missing values, use `aov` with `Error` term do describe the structure of the spatial pseudoreplication. See Chapter 11.4 for example. 

2. If the experiment is not balanced, need to use `lme` or `lmer` for model simplication to estimate the p values of the significant interaction terms. 


```{r}
# Linear model for a split-plot experiment 

yields <- read.table("splityield.txt", header = TRUE)
attach(yields)
head(yields)

library(nlme)

model <- lme(yield ~ irrigation * density * fertilizer, 
             random = ~ 1 | block/irrigation/density)
summary(model) # the only significant effect is irrigation 

# simplify the model 
model <- lme(yield ~ (irrigation + density + fertilizer)^2, 
             random = ~ 1 |  block/irrigation/density)
summary(model)

# remove the fertilizer by density interaction 
model <- lme(yield ~ irrigation * density + irrigation * fertilizer, 
             random = ~ 1 | block/irrigation/density)
summary(model)


# use anova to compare models 

model.lme <- lme(yield ~ irrigation * density * fertilizer,
                  random = ~ 1| block/irrigation/density, method = "ML")

model.lme.2 <- update(model.lme,~. - irrigation:density:fertilizer)
anova(model.lme, model.lme.2)


# remove two way interaction 
model.lme.3 <- update(model.lme.2, ~. - density:fertilizer)
anova(model.lme.3, model.lme.2)


model.lme.4 <- update(model.lme.3,~. - irrigation:fertilizer)
anova(model.lme.3, model.lme.4) # model3 is better 

model.lme.5 <- update(model.lme.2, ~. - irrigation:density)
anova(model.lme.5, model.lme.2)

summary(model.lme.3)

plot(model.lme.3)

plot(model.lme.3, yield ~ fitted(.))

qqnorm(model.lme.3, ~ resid(.) | block) # close to normal distribution 


# do the analysis by lmer 
library(lme4)
b <- block
bi <- block:irrigation

bid <- block:irrigation:density

model1 <- lmer(yield ~ irrigation * density * fertilizer 
               + (1|b) + (1|bi) + (1|bid), REML = FALSE)
print(model1, cor = F) # switch off the matrix of correlations for the fixed effects 


detach(yields)

``` 

## Mixed-effects models with temporal pseudoreplication 

When the random effect(for example, week) is continuous, we should use `$random = \tilde{} week | plant$ instead of $random = \tilde{} 1 | week$, while the latter used for categorical random effects. 

```{r}

results <- read.table("fertilizer.txt", header = TRUE)
attach(results)
head(results)
# root is y 


library(nlme)
library(lattice)

# trellis plotting, 
# convert dataframe into groupedData object, 
# specify the nesting structure 
# indicate the fixed effect by defining "fertilizer" as outer to this nesting(fixed effect)

# week is random effects 
results <- groupedData(root ~ week | plant, outer = ~ fertilizer, results)
str(results)

plot(results)

plot(results, outer = TRUE)

# linear mixed effects using REML
model <- lme(root ~ fertilizer, random = ~week|plant)
summary(model)

# one-way anova on non-pseudoreplicated data 
model2 <- aov(root ~ fertilizer, subset =(week == 10))
summary(model2)

summary.lm(model2)

# the above two models are slightly different
# lm/aov estimates linear models by maximum likelihood estimates of the parameters based on arithmetic means 
# lme use BLUP estimates 

detach(results)
``` 


## Times series analysis in mixed-effects models 

* Use `ACF` to check the autocorrelation structure of residuals. 

* Model autocorrelation structure with standard `corStruct` classes, see Chapter 26.6 for more details(page 863).


```{r}

data(Ovary)
attach(Ovary)
names(Ovary)
str(Ovary)
# follicles is y 

plot(Ovary)

# fit model based on prior knowledge and estimate the residuals correlation structure 
model <- lme(follicles ~ sin(2 * pi * Time) + cos(2 * pi * Time),
             data = Ovary, random = ~ 1| Mare)
summary(model)

plot(ACF(model), alpha = 0.05) # alpha = 0.05 shows the 95% critical lines 

# assume the first two lags have non zero correlations 
# moving average 
model2 <- update(model, correlation = corARMA(q = 2))
anova(model, model2) # better 


# fit first order autoregressive model 
model3 <- update(model2, correlation = corAR1())
anova(model2, model3) # better than model2 


# error checking for model3 
plot(model3)

# plot by mare 
plot(model3, resid(., type = "p") ~ fitted(.) | Mare) # OK 

# check normal error assumption 
qqnorm(model3, ~ resid(.) | Mare) # OK 
detach(Ovary)

```


## Random effects in designed experiments 

Re-analyze the data in Chapter 11.4, instead of using `aov` and `Error`, we specify linear mixed effects model here. 

```{r}
dd <- read.table("rats.txt", header = TRUE)
attach(dd)
head(dd)
# Glycogen is y 
str(dd)

# convert covariates into factors 
Treatment <- factor(Treatment)
Liver <- factor(Liver)
Rat <- factor(Rat)

# unique factor levels for each rat and each liver bit 
rat <- Treatment:Rat
rat
str(rat)
liver <- Treatment:Rat:Liver
liver 


# fit the model 
library(lme4)
model <- lmer(Glycogen ~ Treatment + (1 | rat) + (1 | liver))
summary(model)


# express variance components in percentages 
vars <- c(14.167, 36.065, 21.167)
100 * vars/sum(vars)
# 19.8% is between liver bits within rats 
# 29.6% between readings within liver bits within rats 
detach(dd)
``` 


## Regression in mixed effects models 

1. Use `lmList` to fit lots of linear regression models 

2. Use `lme` to fit one mixed-effects model 



```{r}

yields <- read.table("farms.txt", header = TRUE) 
attach(yields)

names(yields)
str(yields)
# size is y 
table(farm)

plot(N, size, pch = rep(16:18, each = 40), col = farm)


# fit  a set of linear models 
linear.models <- lmList(size ~ N | farm, data = yields)

coef(linear.models) # intercepts and slopes vary a lot 


# fit mixed effects model specified entirely in terms of random effects 
library(nlme)
random.model <- lme(size ~ 1, random = ~ N | farm)
coef(random.model) # less extreme 


# plot the intercepts and slopes from the two models 
mm <- coef(random.model)
ll <- coef(linear.models)

par(mfrow=c(1,2))
plot(ll[, 1], mm[, 1], pch = 16, xlab = "linear",
     ylab = "random effects", main = "Intercept")
abline(0, 1)
plot(ll[, 2], mm[, 2], pch = 16, xlab = "linear",
     ylab = "random effects", main = "Slope")
abline(0,1)
par(mfrow = c(1, 1))


# fit mixed model with both fixed effects and random effects 
# use method = "ML" for model comparisons 

farm <- factor(farm)
mixed.model1 <- lme(size ~ N * farm, random = ~ 1 | farm, method = "ML")
mixed.model2 <- lme(size ~ N + farm, random = ~ 1 | farm, method = "ML")
mixed.model3 <- lme(size ~ N, random = ~ 1 | farm, method = "ML")
mixed.model4 <- lme(size ~ 1, random = ~ 1 | farm, method = "ML")
anova(mixed.model1, mixed.model2, mixed.model3, mixed.model4)
# model2 is selected 


# do analysis of variance 
model <- lm(size ~ N * factor(farm))
summary(model)




model2 <- lm(size ~ N + factor(farm))
anova(model, model2) # model2 selected 

model3 <- lm(size ~ N)
anova(model2, model3)

detach(yields)
```


## Generalized linear mixed models 

* `glmer` from `lme4` package is used. 

* Average the random effects and only work on the fixed effects, if it's poisson error, use code like: 

 *d2<-aggregate(data,list(farm,field),mean)*
 
 *model<-lm(log(count)~factor(farm)+factor(field),data=d2)*
 
 *summary(model)*




# Chapter 20 Non-linear Regression 

When the relationship between y and x cannot be linearized by transformation of the response variable or(and) explanatory variables, non-linear regression will be useful. 

`nls` stands for non-linear least squares. 

**Frequently used non-linear functions**: 

Name|Equation 
------------|--------------------
**Asymptotic functions** | 
Michaelise-Menten | $y = \frac{ax}{1+bx}$
2-parameter asymptotic exponential | $y = a(1-e^{-bx})$
3-parameter asymptotic exponential | $y = a - be^{-cx}$
**S-shaped functions**|
2-parameter logistic | $y = \frac{e^{a+bx}}{1+e^{a+bx}}$
3-parameter logistic| $y = \frac{a}{1+be^{-cx}}$
4-parameter logistic | $y = a + \frac{b-a}{1+e^{(c-x)/d}}$
Weibull | $y = a -be^{-cx^d}$
Gompertz | $y = ae^{-be^{-cx}}$
**Humped curves** |
Ricker curve | $y = axe^{-bx}$
First-order compartment | $y = ke^{-e^ax} - e^{-e^bx}$
Bell-shaped | $y =ae^{-\|bx\|^2}$
Biexponential | $y = ae^{bx} - c e^{-dx}$


```{r}
deer <- read.table("jaws.txt", header = TRUE)
attach(deer)
head(deer)
# bone is y 

plot(age, bone, pch = 21, col = "purple", bg = "green")

# fit 3 parameter asymptotic exponential 
model <- nls(bone ~ a - b * exp( - c * age), start = list(a = 120, b = 110, c = 0.064))
summary(model)


# 2 parameter 
model2 <- nls(bone ~ a * (1 - exp(- c * age)), start = list(a = 120, c = 0.064))
anova(model, model2) # minimal adequate 

# add fitted lines 
av <- seq(0, 50, 0.1)

bv <- predict(model2, list(age = av))
lines(av, bv, col = "red")

summary(model2)
str(summary(model2))

sum.model2 <- summary(model2)
sum.model2$sigma
sum.model2$df[2]

# sum of squares of error for model2 
sse <- as.vector((sum.model2$sigma)^2 * sum.model2$df[2])
sse


# total variation 
null <- lm(bone ~ 1)
str(summary.aov(null)) # one list 

sst <- as.vector(unlist(summary.aov(null)[[1]][2]))
sst


# percentage of variation explained by the model 
100*(sst - sse)/sst




# compare Michaelis-Menten and asyptotic exponential 
(model3 <- nls(bone ~ a * age/(1 + b * age), start = list(a = 8, b = 0.08)))

# add fitted lines to the plot 
yv <- predict(model3, list(age = av))
lines(av, yv, col = "blue")
legend("topleft", legend = c("Michaelis-Menten", "2-parameter asyptotic"),
       col = c("blue", "red"), lty = 1)

detach(deer) 
``` 

## Generalized additive models 

When we don't have any theory or any mechanistic model to suggest a particular functional form to describe the relationship, GAM will be useful. 

```{r}
rm(x, y)
humped <- read.table("hump.txt", header = TRUE)
attach(humped)
names(humped)
plot(x, y, pch = 21, col = "blue", bg = "lavender") 

library(mgcv)

model <- gam(y ~ s(x))

xv <- seq(0.5, 1.3, 0.01)
yv <- predict(model, list(x = xv))
lines(xv, yv)

summary(model)

detach(humped)
``` 


## Grouped data for non-linear estimation 

* `nlsList` fits the same functional form of a group of subjects by the "|" 

* `nlme` fits the nonlinear mixed effects model 

More details on <https://dnett.github.io/S510/29GLMMannotated.PDF>. 


```{r}
reaction <- read.table("reaction.txt", header  = TRUE)
attach(reaction)
head(reaction)
# rate is y 


plot(enzyme, rate, pch = 20 + as.numeric(strain), bg = 1+as.numeric(strain))

library(nlme)

# fit the same model but with different parameters for each strain 
model <- nlsList(rate ~ c + a * enzyme/(1 + b * enzyme)|strain,
                  data = reaction, start = c(a = 20, b = 0.25, c = 10))


summary(model)


# plot 
reaction <- groupedData(rate ~ enzyme | strain, data = reaction)
plot(reaction)

# fit non-linear mixed effects model 
model2 <- nlme(rate ~ c + a * enzyme/(1 + b * enzyme), fixed = a + b + c ~ 1,
      random = a + b + c ~ 1 | strain, data = reaction, start = c(a = 20, b = 0.25, c = 10))

plot(augPred(model2)) 
# augPred returns a data frame with four columns representing, respectively, 
# the values of the primary covariate, the groups (if object does not have a 
# grouping structure, all elements will be 1), the predicted or observed values, 
# and the type of value in the third column
model2
summary(model2)

coef(model2)
v <- vcov(model2)
v
plot(enzyme, rate, pch = 20 + as.numeric(strain), bg = 1 + as.numeric(strain))

xv <- seq(min(enzyme), max(enzyme), length = 100)
for(i in 1:5){
yv <- coef(model)[i, 3] + coef(model)[i, 1] * xv/(1 + coef(model)[i, 2] * xv) 
lines(xv, yv, col = (i + 1)) }

detach(reaction)
rm(xv, yv)

``` 


## Non-linear time series models (temporal pseudo-replication) 


```{r}

nl.ts <- read.table("nonlinear.txt", header = TRUE) 
attach(nl.ts)
head(nl.ts)

# group by dish 
growth <- groupedData(diam ~ time | dish, data = nl.ts) 

model <- nlme(diam ~ a + b * time/(1 + c * time),
      fixed = a + b + c ~ 1,
      random = a + b + c ~ 1,
      data = growth,
      correlation = corAR1(),
      start = c(a = 0.5, b = 5, c = 0.5))
summary(model)

coef(model)

plot(augPred(model))



detach(nl.ts)
``` 


## Self-starting functions 

Function | Description 
-------------|--------------------------------------------------------------------------------
SSasymp | asymptotic regression model;
SSasympOff | asymptotic regression model with an offset; 
SSasympOrig | asymptotic regression model through the origin;
SSbiexp | biexponential model;
SSfol | first-order compartment model;
SSfpl | four-parameter logistic model;
SSgompertz | Gompertz growth model;
SSlogis | logistic model;
SSmicmen |  Michaelis–Menten model;
SSweibull | Weibull growth curve model.



```{r}

# self starting Michaelis-menten model 
data <- read.table("mm.txt", header = TRUE)
attach(data)
names(data)

plot(rate ~ conc, pch = 16)

model <- nls(rate ~ SSmicmen(conc, a, b)) 
# a is the max value of rate, b is the value of conc at which half of the max response attained 
summary(model)

xv <- seq(0, 1.2, 0.01)
yv <- predict(model, list(conc = xv))
lines(xv, yv, col = "blue")
detach(data)


# self starting asymptotic exponential model 
deer <- read.table("jaws.txt", header = TRUE)
attach(deer)
names(deer)

model <- nls(bone ~ SSasymp(age, a, b, c))
plot(age, bone, pch = 16)
xv <- seq(0, 50, 0.2)
yv <- predict(model, list(age =xv))
lines(xv, yv)
summary(model)

detach(deer)



# self starting logistic 
sslogistic <- read.table("sslogistic.txt", header = TRUE)
attach(sslogistic)

names(sslogistic)

plot(density ~ log(concentration), pch = 16, col = "green3")

model <- nls(density ~ SSlogis(log(concentration), a, b, c))

xv <- seq(-3, 3, 0.1)
yv <- predict(model, list(concentration = exp(xv)))
lines(xv, yv, col = "red")
summary(model)
detach(sslogistic)





# four-parameter logistic 
data <- read.table("chicks.txt", header = TRUE) 
attach(data)
names(data)

model <- nls(weight ~ SSfpl(Time, a, b, c, d))
xv <- seq(0, 22, 0.2)
yv <- predict(model, list(Time = xv))
plot(weight ~ Time, pch = 21, col = "red", bg = "green4")
lines(xv, yv, col = "navy")
summary(model)
detach(data)



# self start weibull growth function 
weights <- read.table("weibull.growth.txt", header = TRUE) 
attach(weights)
names(weights)


model <- nls(weight ~ SSweibull(time, Asym, Drop, lrc, pwr))
summary(model)

xt <- seq(2, 22, 0.1)
yw <- predict(model, list(time = xt))
plot(time, weight, pch = 21, col = "blue", bg = "orange")
lines(xt, yw, col = "blue2")
detach(weights)



# self starting first order compartment function 
foldat <- read.table("fol.txt", header = TRUE)
attach(foldat)
names(foldat)

model <- nls(conc ~ SSfol(Dose, Time, a, b, c))
summary(model)

xv <- seq(0, 25, 0.1)
yv <- predict(model, list(Time = xv))
plot(conc ~ Time, pch = 21, col = "blue", bg = "red")
lines(xv, yv, col = "green4")

detach(foldat)
``` 


## Bootstrapping a family of non-linear regression 


```{r}
library(MASS)
data(stormer)
attach(stormer)

model <- nls(Time ~ b * Viscosity/(Wt - c), start = list(b = 29, c = 2))
summary(model)

plot(Viscosity, Time, pch = 16, col = 1 + as.numeric(factor(Wt)))
xv <- 0:300
yv <- predict(model, list(Wt = 20, Viscosity = xv))
lines(xv, yv, col = 2)
yv <- predict(model, list(Wt = 50, Viscosity = xv))
lines(xv, yv, col = 3)
yv <- predict(model, list(Wt = 100, Viscosity = xv))
lines(xv, yv, col = 4)


# homemade function to do bootstrap 
bv <- numeric(1000)
cv <- numeric(1000)
for(i in 1:1000){
ss <- sample(1:23, replace = TRUE) 
y <- Time[ss]
x1 <- Viscosity[ss]
x2 <- Wt[ss]
model <- nls(y ~ b * x1/(x2 - c), start = list(b = 29, c = 2))
bv[i] <- coef(model)[1]
cv[i] <- coef(model)[2] 
}

quantile(bv, c(0.025, 0.975))

quantile(cv, c(0.025, 0.975))




# bootstrap by boot package 
library(boot)

rs <- resid(model)
fit <- fitted(model)

storm <- data.frame(fit, Viscosity, Wt)

statistic <- function(rs, i){
storm$y <- storm$fit + rs[i]
coef(nls(y ~ b * Viscosity/(Wt - c), data = storm, start = coef(model)))
}
# note: some iterations may have singular gradient and thus error message, re-run the program in this case 
boot.model <- boot(rs, statistic, R = 1000)
boot.model

boot.ci(boot.model,index=1)

boot.ci(boot.model,index=2)

detach(stormer)

``` 
